---
title: "gwas2crispr: From GWAS to CRISPR-ready Files"
pagetitle: "gwas2crispr: From GWAS to CRISPR-ready Files"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{gwas2crispr: From GWAS to CRISPR-ready Files}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

## Overview

`gwas2crispr` prepares genome-wide association study (GWAS) results for downstream clustered regularly interspaced short palindromic repeats (CRISPR) workflows.

The package retrieves significant single-nucleotide polymorphisms (SNPs) for an Experimental Factor Ontology (EFO) trait from the EMBL-EBI GWAS Catalog REST API v2 and returns CRISPR-ready outputs for the GRCh38/hg38 human genome build.

The main outputs are:

* comma-separated values (CSV) tables,
* Browser Extensible Data (BED) files,
* optional FASTA sequence files.

## Installation

Install from CRAN:

```{r}
install.packages("gwas2crispr")
```

Optional packages for FASTA output:

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "Biostrings",
  "GenomeInfoDb",
  "BSgenome.Hsapiens.UCSC.hg38"
))
```

Development version:

```{r}
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")

devtools::install_github("leopard0ly/gwas2crispr")
```

## Fetch GWAS associations

```{r}
library(gwas2crispr)

gwas_data <- fetch_gwas(
  efo_id  = "EFO_0000707",
  p_cut   = 1e-6,
  verbose = FALSE
)

names(gwas_data)
head(gwas_data$associations)
```

## Run without writing files

By default, no files are written.

```{r}
res <- run_gwas2crispr(
  efo_id     = "EFO_0000707",
  p_cut      = 1e-6,
  flank_bp   = 300,
  out_prefix = NULL,
  verbose    = FALSE
)

res$summary
head(res$snps_full)
head(res$bed)
```

## Write files safely

To write output files, provide `out_prefix`. In examples, use `tempdir()`.

```{r}
out_prefix <- file.path(tempdir(), "lung")

res <- run_gwas2crispr(
  efo_id     = "EFO_0000707",
  p_cut      = 1e-6,
  flank_bp   = 300,
  out_prefix = out_prefix,
  verbose    = FALSE
)

res$written
```

Expected output paths:

```{r}
paste0(out_prefix, "_snps_full.csv")
paste0(out_prefix, "_snps_hg38.bed")
paste0(out_prefix, "_snps_flank300.fa")
```

The FASTA file is created only when the optional genome packages are available.

## Output structure

```{r}
names(res)
```

Common outputs:

```{r}
res$summary
res$snps_full
res$bed
res$fasta
res$written
```

## Session information

```{r, eval=TRUE}
sessionInfo()
```
